---
title: "Some Commonly Employed Practices in Data Wrangling"
subtitle: "Urbana Crime Incidents Data"
author: "Dr. Weijia Jia"
date: "2024-03-21"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```


## Example: Urbana Police Incidents from 2010 - Feb 2023

Data Source: Urbana's Open Data website 

[Urbana Police Incidents Dataset Link](https://data.urbanaillinois.us/Police/Police-Incidents-from-1988-Feb-2023/uj4k-8xe8/about_data)

Police incidents from 1988 until February 2023, when the Urbana Police began using a new computer system. Records before 1988 are incomplete - and could be data entry errors. | MAPPING NOTE! Some addresses will not appear on maps because streets no longer exist or are not recognized by the mapping databases.

In this workshop, we only use the data from 2010 to 2023.

### Read in the csv datasets

```{r}
library(tidyverse)
df1 <- read_csv("Urbana Police Incidents Data (2010-2014).csv")
df2 <- read_csv("Urbana Police Incidents Data (2015-2019).csv")
df3 <- read_csv("Urbana Police Incidents Data (2020-2023).csv")


```



## Merge the data vertically or horizontally

### Stack the datasets -- merge vertically
```{r}
### Address the discrepancy in data types for WEAPONS.CODE.1 between df1 and df3.
df3$WEAPONS.CODE.1 <- as.character(df3$WEAPONS.CODE.1)

### Stack three data frames
UC <- bind_rows(df1, df2, df3)
```


### Merge the dataset horizontally `bind_cols()`

## Slice some data for preliminary research

Preliminary exploration -- extract a sample of the data for easier management.

```{r}
UC.sample <- UC %>% slice_sample(n = 1000) 
```


## How many missing values in each column?

```{r}
UC %>% 
  summarize_all(~sum(is.na(.))) %>% 
  print(width=Inf) ### Show all columns
```

## Distinct groups

## List the distinct groups

```{r}
UC %>% 
  distinct(CRIME.CATEGORY)
```


## Find the number of distinct groups

```{r}
UC %>% 
  select(CRIME.CATEGORY) %>% 
  n_distinct()
```

## Deal with time and date

```{r}
UC <- UC %>% 
  mutate(Year.occ = year(mdy(DATE.OCCURRED)),
         Month.occ = month(mdy(DATE.OCCURRED)),
         Day.occ = day(mdy(DATE.OCCURRED)),
         Hour.occ = hour(hms(TIME.OCCURRED)))

### Check if Year.occ and YEAR.OCCURRED are identical
identical(UC$Year.occ, UC$YEAR.OCCURRED)
# Note that, some non-identical case are due to different data type

### Discover where the difference lies
setdiff(UC$Year.occ, UC$YEAR.OCCURRED)
```


## Count number in each category, sort in order

```{r}
UC %>% 
  count(CRIME.DESCRIPTION) %>% 
  arrange(desc(n)) %>% ## sort in descending order
  head() ## Show the first 6 categories only
```

## Combine some categories in a variable

```{r}
UC %>% 
  mutate(
    YEAR_before_19 = case_when(
      YEAR.OCCURRED <2019 ~ "Yes",
      TRUE ~ "No"
    )
  ) %>% 
  count(YEAR_before_19)
```


## Check if two variables are one-on-one correspondence

```{r}

m<- UC %>% 
  distinct(CRIME.CATEGORY, CRIME.CATEGORY.DESCRIPTION) %>% 
  table()

all(rowSums(m)==1) && all(colSums(m)==1)
```


## geom_bar() vs geom_col()

```{r}
UC<- UC %>% 
  mutate(
    YEAR_before_19 = case_when(
      YEAR.OCCURRED <2019 ~ "Yes",
      TRUE ~ "No"
    ))

# geom_bar() <- pre-count not needed
UC %>% 
  ggplot(aes(YEAR_before_19)) + 
  geom_bar()

# geom_col() <- pre-count needed
UC %>% 
  count(YEAR_before_19) %>% 
  ggplot(aes(x = YEAR_before_19, y = n))+
  geom_col()
```


### color or fill, in aes() or not

```{r}
UC<- UC %>% 
  mutate(
    Traffic_Offenses = case_when(
      CRIME.CATEGORY.DESCRIPTION == "Traffic Offenses" ~ "Yes",
      TRUE ~ "No"
    ))
```

```{r}
### color vs fill

UC %>% 
  ggplot(aes(x=Traffic_Offenses, color=YEAR_before_19))+
  geom_bar(position="dodge")

UC %>% 
  ggplot(aes(x=Traffic_Offenses, fill=YEAR_before_19))+
  geom_bar(position="dodge")
```

```{r}
### Color #13294B is UIUC navy blue, #E84A27 is UIUC orange
UC %>% 
  ggplot(aes(x=Traffic_Offenses, fill="#13294B"))+
  geom_bar()

### The color doesn't looks like navy blue. What's wrong?
UC %>% 
  ggplot(aes(x=Traffic_Offenses))+
  geom_bar(fill="#13294B", color="#E84A27", linewidth=2)
```


## Seperate strings in one column and create barplot based on frequency

```{r, out.width="120%"}
UC %>% 
  mutate(
    CRIME.CATEGORY.DESCRIPTION1 = str_split(CRIME.CATEGORY.DESCRIPTION, "/")
  )  %>% 
  unnest(CRIME.CATEGORY.DESCRIPTION1) %>% 
  count(CRIME.CATEGORY.DESCRIPTION1) %>%
  ggplot(aes(x=reorder(CRIME.CATEGORY.DESCRIPTION1, n), y = n))+ # arrange bars in order
  geom_col()+
  coord_flip()+ 
  theme(axis.text = element_text(size = 5))
```

